---
title: "tvvar: Sparse Estimation of Time-Varying VAR Models"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{tvvar: Sparse Estimation of Time-Varying VAR Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installation

The package can be installed with

```{r, eval=FALSE}
install.packages("tvvar")
```

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tvvar)
set.seed(1)
```

Make sure to also install `apg`, the package for `accelerated proximal gradient descent` from https://github.com/jpvert/apg:

```{r, eval=FALSE}
install_github("jpvert/apg", force=TRUE)
```

## Background

This package builds on dynamic factor VAR models with conditional heteroskedastic errors [@gorgi2019dfvar]. 

[TODO]: Elaborate and refer to own paper

## Quick start
An example of a simulated dataset with `N` = 2, `r` = `p` = 1, with which can be called by:

```{r}
data(simdata, package = "tvvar")  
str(simdata$Y[1:5, ])
```

gives a list with the following components: 
- `Y`: a T × N matrix containing the observed time series data.\
- `factors`: a T × r matrix of the latent factor time series.\
- `shocks`: a T × N matrix of the VAR shocks (innovations).

We run the unpenalized estimation using ML on the simulated dataset and retrieve the estimated parameters and log-likelihood:

```{r, message=FALSE, warning=FALSE}
fit_ml <- unpenalized_estimate(
  data             = simdata$Y,
  p                = 1,
  r                = 1,
  zero.mean        = TRUE,
  phi_f_structure  = matrix(1, 2, 2),
  method           = "ML",
  init             = "default")
```

[TODO]: Make summary more compact

And run a summary.......

```{r}
tvvar_summary(fit_ml)
```

## Unpenalized estimation
The function `unpenalized_estimate()` lets the user specifies the data as input, given as a T x N matrix. The `unpenalized_estimate()` function returns a `S3` class object `tvvar` which contains the estimated parameters and scores on fit of the model. The function takes the following arguments:

```{r, echo = TRUE, eval = FALSE}
unpenalized_estimate(
  data,                       
  p = 1,                      
  r = 1,                      
  zero.mean = TRUE,           
  phi_f_structure,            
  method = c("ML","EM"),    
  control    = list(maxit = 2000, trace = 0),           
  em_control = list(max_iter = 200, tol = 1e-3, trace = TRUE), 
  init = c("default","random","custom"),              
  init_list = NULL                                     
)
```

### Data
A **T × N** numeric matrix containing the observed time series data.  
Each column corresponds to a variable, and each row to a time observation.  
The function assumes the data are ordered chronologically (oldest to newest).

### Lag order `p`
Integer specifying the **lag order** of the VAR component.  
This determines how many past observations enter the autoregressive structure.

### Number of factors `r`
Integer specifying the **number of latent factors** driving the time variation in coefficients.  
Typically small (e.g., 1–3) relative to the number of observed variables.

### Intercept `zero.mean`
Logical; if `TRUE`, the intercept is fixed at zero and the data should be **demeaned** beforehand.  
If `FALSE`, an intercept is estimated in the constant block `Φ_c`.

### Factor loadings structure `phi_f_structure` 
Specifies which elements of `Φ_f` (the factor loadings) are free or fixed to zero.  
Can be provided as:
- a 3D array `[N, N, r]`,  
- a list of `r` `N × N` matrices, or  
- a block matrix of size `N × (N * r)`.  
Entries set to zero are treated as fixed.

### Optimization method `method`
Character; either `"ML"` for **maximum likelihood estimation**  
(using numerical optimization) or `"EM"` for the **expectation–maximization algorithm**.  
Both methods estimate the same parameters, but `EM` can be more stable for latent-factor models.

### Control settings for ML `control`
List of optimizer settings for `"ML"` estimation, e.g.  
`list(maxit = 2000, trace = 0)`.  
These are passed directly to `optim()`.

### Control settings for EM `em_control`
List of settings for `"EM"` estimation, e.g.  
`list(max_iter = 200, tol = 1e-3, trace = TRUE)`.  
Controls the maximum number of EM iterations, tolerance, and verbosity.

### Initialization type `init`
Initialization mode for the parameter vector.  
Accepts `"default"`, `"random"`, or `"custom"`:
- `"default"` — deterministic starting values based on simple OLS or heuristics.  
- `"random"` — random but stationary initialization (useful for robustness checks).  
- `"custom"` — user-specified starting values via `init_list`.

### Initialization values `init_list`
Optional **named list** supplying custom starting values.  
The user can override any subset of:
`A`, `B`, `phi_r`, `Omega`, `Phi_c`, or `Phi_f`.  
Entries not provided are filled with defaults.  
Note that `Phi_f` should include only the **free entries** (those set to 1 in `phi_f_structure`).


### Sparse data set
In many macro applications, only a few VAR coefficients truly vary over time. We demonstrate how penalization improves fit and structure recovery when the true factor-loadings block \( \Phi_f \) is **sparse**. Using `simulate_tvvar_data()`, 

```{r}
N <- 2; p <- 2; r <- 2
Phi.f <- matrix(0, nrow = N, ncol = N * p * r)
Phi.f[1, 1] <-  0.25  
Phi.f[2, 4] <- -0.20  
Phi.f[1, 7] <-  0.15  

sim_sparse <- simulate_tvvar_data(
  T = 1000, N = N, p = p, r = r,
  A2 = 0.1, B2 = 0.75,
  pphi = rep(0.95, r),
  omega = matrix(c(0.30, 0.20, 0.20, 0.30), N, N, byrow = TRUE),
  Phi.c = matrix(c(
  0.30,  0.10,   0.00,  0.10,  
  0.00,  0.25,   0.10,  0.00
  ), nrow = N, byrow = TRUE),
  Phi.f = Phi.f
)
```

## Real data example
We also have the real dataset of US macroeconomics, which can be loaded by:

```{r, eval=TRUE, echo=TRUE}
data(macro_data)
str(macro_data[1:5, ])
```

This dataset contains 5 variables: date, unemployment rate, inflation rate and federal funds rate, from 1960 to 2022 (T = 747). The data is quarterly and seasonally adjusted. The data is stored in a T x N matrix where each column corresponds to a variable.

[TODO]: Apply unpen/pen.